---
title: "Lab 1B"
name: ""
output: pdf_document
---

The following dataset, titled "Rain in Australia," was obtained via Kaggle.com via author Joe Young et. al.

This dataset contains daily weather observations from numerous Australian weather stations during the year 2016.  The dataset measured variables: Date, Location, MinTemp (C), MaxTemp (C), Rainfall (mm), Evaporation (mm per 24 hours), Sunshine (hours per day), WindGustDir (strongest gust in 24 hours), WindGustSpeed (km/h), WindDir9am (direction at 9 AM), ..., RainToday (Yes/No) and RainTomorrow (Yes/No).

```{r}
rain = read.csv("weatherAUS.csv", sep=",", header=TRUE)
rain = na.omit(rain)
head(rain, 4)
```



## Question 1
Using 1000 randomly sampled observations from the dataset `rain`, plot a scatter plot of `MaxTempF` (x) vs. `Humidity3PM` (y).  `MaxTempF` should be the `MaxTemp` variable converted to degrees F.  Color the point on the scatterplot blue if it rained that day and red if it did not rain that day.  The plot should have x- and y- axis labels in addition to a title.  After the plot, include a discussion on observations from the scatterplot.

Hints:
- To sample from `rain`, use `sample_n()` from package `dplyr`.
- To create MaxTempF, use conversion (9/5)C + 32 to change C to F.
- Create a second new column named Colors that assigns a color based on RainToday variable.

```{r}
library(dplyr)

```



## Question 2

### Part A
Plot a histogram of `Rainfall`, including all values from `rain` dataset given that it had rained that day.  Display density on the y-axis.  Label the axes and give the graph a title.  Colorize the bars.  Include a discussion after the plot.

```{r}

```


### Part B
Find the mean and median of the `Rainfall` in mm for days that it rained in Australia.  Do these numbers correspond with the image in part A?

```{r}

```



## Question 3

### Part A
Plot a bar graph representing *total* `Rainfall` in 2016 by `Location`.  Make sure to label the x- and y- axes, colorize the bars, include a title, and include the appropriate response labeled above each bar.  Utilize all observations from dataset `rain`. Include a discussion after the plot.

```{r}

```


### Part B
Plot a bar graph representing *average* `Rainfall` per day based on `Location`.  Make sure to label the x- and y- axes, colorize the bars, include a title, and include the appropriate response labeled above each bar.  Utilize all observations from dataset `rain`. Include a discussion after the plot.

```{r}

```


### Part C
Plot a bar graph representing *total* count of days rained in 2016 based on `Location`.  Make sure to label the x- and y- axes, colorize the bars, include a title, and include the appropriate response labeled above each bar.  Utilize all observations from dataset `rain`. Include a discussion after the plot.

```{r}

```



## Question 4

For locations `Brisbane`, `Melbourne`, `Perth`, and `Sydney`, plot side by side boxplots of the variable `Rainfall` for days where it rained.  The graph should include a title, x- and y-axis labels, and unique colors for the 4 boxplots.  The data should include all observations from the `rain` dataset.  Include a discussion after the plot.

```{r}

```
